Automated discovery of interpretable gravitational-wave population models

### 2022-07-11 src/tex/ms.tex

\begin{abstract}
We present an automatic approach to discover analytic population models for gravitational-wave (GW) events from data.
As more gravitational-wave (GW) events are detected, flexible models such as Gaussian Mixture Models have become more important in fitting the distribution of GW properties due to their expressivity.
However, flexible models come with many parameters that lack physical motivation, making interpreting the implication of these models challenging.
In this work, we demonstrate the use of symbolic regression to distill such flexible models into interpretable analytic expressions.
We recover common GW population models such as a power-law-plus-Gaussian, and find a new empirical population model which combines accuracy and simplicity.
This demonstrates a promising pathway to discover interpretable population models, automatically, from the GW merger catalogs and other astrophysical phenomena.
\end{abstract}

\section{Introduction}
\begin{figure*}[hbt!]
% \includegraphics[width=\textwidth]{figures/GWTC3_m1_combined.pdf}
\includegraphics[width=\textwidth]{figures/Page2.pdf}
\caption{An illustration of our proposed pipeline, which first
(\textit{far left})
requires observations of individual source properties (here, events shown from GWTC3);
then
(\textit{middle left}) 
infers the deconvolved population posterior density;
followed by (\textit{middle right})
using symbolic regression to search for an analytic model which approximates the density, using
operators common to existing population models; and, finally
(\textit{far right}) distilling physical insights
from the recovered population model.
}
%\caption{\textit{left}: posterior density of the primary mass from all events reported in GWTC3. Each line is the normalized posterior density for one event.
%\textit{middle}: posterior predictive distribution of the Gaussian mixture model reported in \cite{LIGOScientific:2021psn}.
%\textit{right}: equations tabulated in table \ref{tab:equations}}
\label{fig:Gaussian_m1_fit}
\end{figure*}
The rate of gravitational wave (GW) detections is increasing exponentially as the sensitivity of GW detectors improves \cite{LIGOScientific:2021djp} since the discovery in 2015 \cite{Abbott:2016blz}.
Today there exist 79 official events, the majority of them being Binary Black Holes (BBHs).
As more GW events are observed, more features in the distribution of GW events become resolved and available for discovery.
Currently, one of the most common ways to understand the observed GW population is through phenomenological modeling.
For example, to understand the distribution of primary mass in observed BBHs, one would hand-craft a model of the primary mass distribution, such as a power-law distribution \cite{LIGOScientific:2021psn}.
By comparing the model to the data, one can measure the parameters which characterize the model, such as the spectral index. \todo{Could make a statement about how power laws typically have some astrophysical meaning - hence a good operator, and why they are used in population models.}

@@

The lowest complexity (16) equation returned is a decaying exponential function with a Gaussian bump at the lower end of the mass distribution.
This agrees with a recent study pointing out there is a very strong excess of events at $m_{1}\sim 10\ M_{\odot}$.
In order to account for the bump around $30\ M_{\odot}$,
the exponential part of this equation has a much higher normalization compared to the others.
This agrees with the analysis done in GWTC3, where a single power law seems to have trouble accounting for the excess of events around $m_{1}\sim 30\ M_{\odot}$\cite{LIGOScientific:2021psn}.

Compared to the lowest complexity equation, the middle complexity (25) equation instead prefers a lower normalization of the exponential function,
but account for the bump around $30\ M_{\odot}$ with a Gaussian component.
This equation is similar to one of the simplest phenomenological models fitted to GWTC3, which is a power law mixed with a Gaussian bump.
The difference between this equation and the one presented in GWTC3 is this equation has an additional bump at $m_{1}\sim 10\ M_{\odot}$,
which is also present in the lower complexity equation.
This suggests the first bump at lower mass has more statistical significance than the second bump.
Although the middle complexity equation is more complex, the loss is about one order of magnitude lower than the next best one with lower complexity (16).
As a result, the score of this equation is higher than the lower complexity equation, making it a favorable equation. This means the second Gaussian component is needed in order to describe the distribution of GWs.

@@

% Brief summary of the paper
In this paper, we use symbolic regression to extract interpretable equations from the \ppd of a flexible model fitted to the catalog of GWs.
We show SR can discover equations which are similar to state-of-the-art phenomenological models.
On top of finding forms that have already been discovered,
our result also shed insights on an open question---whether there exist additional features other than the two prominent peaks in the primary mass distribution \cite{Tiwari:2021yvr}.
Our results suggest any additional features such as tapering of the continuum or wiggles between the two peaks are not statistically significant \todo{I would lighten the language of this part - SR isn’t exactly statistically rigorous}.

Compared to traditional methods such as phenomenological modeling, which relies on an individual's insight,
our pipeline is fully automatic.
This allows the community to discover interpretable patterns in the GW catalog in a data-driven instead of model-driven manner.
Hence, our method is more scalable to more complex and higher-dimensional datasets.

>>>>

\begin{abstract}
We present an automatic approach to discover analytic population models for gravitational-wave (GW) events from data.
As more gravitational-wave (GW) events are detected, flexible models such as Gaussian Mixture Models have become more important in fitting the distribution of GW properties due to their expressivity.
However, flexible models come with many parameters that lack physical motivation, making interpreting the implication of these models challenging.
In this work, we demonstrate symbolic regression can complement flexible models by distilling the posterior predictive distribution of such flexible models into interpretable analytic expressions.
We recover common GW population models such as a power-law-plus-Gaussian, and find a new empirical population model which combines accuracy and simplicity.
This demonstrates a promising pathway to discover interpretable population models, automatically, from the GW merger catalogs and other astrophysical phenomena.
\end{abstract}

\section{Introduction}
\begin{figure*}[hbt!]
% \includegraphics[width=\textwidth]{figures/GWTC3_m1_combined.pdf}
\includegraphics[width=\textwidth]{../static/Gaussian_m1_fit.pdf}
\caption{An illustration of our proposed pipeline, which first
(\textit{far left})
requires observations of individual source properties (here, events shown from GWTC3);
then
(\textit{middle left}) 
infers the deconvolved population posterior density;
followed by (\textit{middle right})
using symbolic regression to search for an analytic model which approximates the density, using
operators common to existing population models; and, finally
(\textit{far right}) distilling physical insights
from the recovered population model.
}
%\caption{\textit{left}: posterior density of the primary mass from all events reported in GWTC3. Each line is the normalized posterior density for one event.
%\textit{middle}: posterior predictive distribution of the Gaussian mixture model reported in \cite{LIGOScientific:2021psn}.
%\textit{right}: equations tabulated in table \ref{tab:equations}}
\label{fig:Gaussian_m1_fit}
\end{figure*}
The rate of gravitational wave (GW) detections is increasing exponentially as the sensitivity of GW detectors improves \cite{LIGOScientific:2021djp} since the discovery in 2015 \cite{Abbott:2016blz}.
Today there exist 79 official events, the majority of them being Binary Black Holes (BBHs).
As more GW events are observed, more features in the distribution of GW events become resolved and available for discovery.
Currently, one of the most common ways to understand the observed GW population is through phenomenological modeling.
For example, to understand the distribution of primary mass in observed BBHs, one would hand-craft a model of the primary mass distribution, such as a power-law distribution \cite{LIGOScientific:2021psn}.
By comparing the model to the data, one can measure the parameters which characterize the model, such as the spectral index. \todo{Could make a statement about how power laws typically have some astrophysical meaning - hence a good operator, and why they are used in population models.}

@@

The lowest complexity (16) equation returned is a decaying exponential function with a Gaussian bump at the lower end of the mass distribution.
This agrees with a recent study pointing out there is a very strong excess of events at $m_{1}\sim 10\ M_{\odot}$,
which corresponds to a smooth turn-on of the black hole mass spectrum \cite{Talbot:2018cva}.
In order to account for the bump around $30\ M_{\odot}$,
the exponential part of this equation has a much higher normalization compared to the others.
This agrees with the analysis done in GWTC3, where a single power law seems to have trouble accounting for the excess of events around $m_{1}\sim 30\ M_{\odot}$\cite{LIGOScientific:2021psn}.

Compared to the lowest complexity equation, the middle complexity (25) equation instead prefers a lower normalization of the exponential function,
but account for the bump around $30\ M_{\odot}$ with a Gaussian component,
which is proposed to be driven by a pile-up from pulsational pair-instability supernovae \cite{Talbot:2018cva}.
This equation is similar to one of the simplest phenomenological models fitted to GWTC3, which is a power law mixed with a Gaussian bump.
The difference between this equation and the one presented in GWTC3 is this equation has an additional bump at $m_{1}\sim 10\ M_{\odot}$,
which is also present in the lower complexity equation.
This suggests the first bump at lower mass has more statistical significance than the second bump.
Although the middle complexity equation is more complex, the loss is about one order of magnitude lower than the next best one with lower complexity (16).
As a result, the score of this equation is higher than the lower complexity equation, making it a favorable equation. This means the second Gaussian component is needed in order to describe the distribution of GWs.

@@

% Brief summary of the paper
In this paper, we use symbolic regression to extract interpretable equations from the \ppd of a flexible model fitted to the catalog of GWs.
We show SR can discover equations which are similar to state-of-the-art phenomenological models.
On top of finding forms that have already been discovered,
our result also shed insights on an open question---whether there exist additional features other than the two prominent peaks in the primary mass distribution \cite{Tiwari:2021yvr}.
Our results suggest interpreting any additional features such as tapering of the continuum or wiggles between the two peaks can be tricky as their significance are unclear. \todo{Clean up language}

The main message of this work is that by combining flexible models with symbolic regression, we can discover interpretable equations that can describe the increasingly complex GW catalog.
Compared to traditional methods such as phenomenological modeling, which relies on an individual's insight,
our pipeline requires minimal human intervene.
Our method is not a model selection method by itself, but a pipeline to discover models for downstream tasks such as model selection and parameter estimation.
Once we obtain a set of equations, we can plug the expressions in a standard HBA pipeline to estimate the parameters of the model and select between models.
This allows the community to discover interpretable patterns in the GW catalog in a data-driven instead of model-driven manner.
In this sense, our method is more scalable to more complex and higher-dimensional datasets.
